An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)

### 2022-05-24 manuscript/main.tex
manuscript updates

\abstract{Matrix reasoning tasks are among the most widely used measures of cognitive ability in the behavioral sciences, but the paucity of matrix reasoning tests in the public domain complicates their use. Here we present an extensive investigation and psychometric validation of the matrix reasoning item bank (MaRs-IB), an open-access set of matrix reasoning items. In a first study, we calibrate the psychometric functioning of the items in the MaRs-IB in a large sample adults participants (N=1501). Using additive multilevel item structure models, we establish that the MaRs-IB has many desirable psychometric properties: its items span a wide range of difficulty, possess medium- to-large levels of discrimination, and exhibit clear associations between item complexity and difficulty. However, we also find that item clones are not always psychometrically equivalent and cannot assumed to be exchangeable. In a second study, we demonstrate how researchers can use the item parameters we estimated to design new MaRs-IB tests using optimal item assembly. With these methods, we design three parallel short-form measures that exhibited good score reliability and convergent validity in a second sample of adults (N=300). We hope that the materials and results made available here will encourage researchers to use the MaRs-IB in their studies.}

@@

Matrix reasoning tasks are among the most widely used measures of cognitive ability in the behavioral sciences. Much of their popularity undoubtedly reflects their versatility of use. Matrix reasoning tasks are strong (albeit impure) indicators of general intelligence \citep{gignac2015raven} and working memory capacity \citep{kane2004generality, unsworth2005working}. They are predictive of important real-world outcomes such as childhood academic achievement \citep{roth2015intelligence} and performance on college entrance exams \citep{frey2004scholastic, koenig2008act}. In low-stakes testing settings (i.e. where a participant incurs little or no cost for poor performance), matrix reasoning tasks can additionally function as measures of motivation, willingness to expend mental effort, and other facets of personality \citep{duckworth2011role, gignac2019maximum}. In studies of psychiatric populations, performance on matrix reasoning tasks have also been used to control for general disruptions to cognitive ability when specific domains of cognition are of primary interest \citep{gillan2016characterizing, rouault2018psychiatric, moutoussis2021decision}.

Unfortunately there are a number of obstacles to using matrix reasoning tasks as part of behavioral research. One challenge is the problem of copyright. Many of the most prominent matrix reasoning tasks, such as the WAIS and WASI matrix reasoning subtests \citep{wechsler1999wechsler, corporation2008wechsler}, are not free to use and have legal restrictions against being computerized. A second challenge is that, across all of the matrix reasoning tests in the public domain, there are relatively few unique items available. The Hagen matrices test \citep{heydasch2014hagen} and ICAR matrix reasoning test \citep{condon2014international}, for example, have only 20 and 16 items respectively. The availability of only a limited number of items raises the possibility of repeat exposure effects, which threaten the validity of these measures  \citep{ng1974applicability, bors2003effect}. This problem is exacerbated in the current era of online experiments, where multiple groups of researchers may be inadvertently administering the same test to the same participants recruited from the same online labor platforms.

Noting these challenges, \cite{knoll2016window} / \cite{chierchia2019matrix} developed and made publicly available the matrix reasoning item bank (MaRs-IB), a collection of 80 matrix reasoning puzzles. Each puzzle in the MaRs-IB consists of a 3x3 matrix containing geometric shapes in eight out of nine cells (Figure \ref{fig:fig00}). Across cells, some or all of these shapes change according to a number of abstract rules. Based on these results, participants must deduce which of four response options correctly completes the matrix. The puzzles in the MaRs-IB vary in their complexity, both in the number of elements per cell and number of rules determining the relations of these elements across cells (Figure \ref{fig:fig00}a/b). To increase the reusability of the MaRs-IB, each puzzle (hereafter referred to as an item template) has six clones. A single item's clones share identical structural characteristics (i.e. number of elements, number of rules) but may vary in their distractors (two unique sets per template) or perceptual features (three unique shape sets per template; Figure \ref{fig:fig00}c/d). Thus, the MaRs-IB addresses several of the issues associated with other matrix reasoning tests.

\cite{chierchia2019matrix} conducted an initial study of the psychometric properties of the MaRs-IB in which a sample of 659 adult, adolescent, and child participants had eight minutes to complete as many MaRs-IB items as possible (items were presented in the same order to all participants). The authors found that performance on this task had good split-half and test-retest reliability. Moreover, they found that participants' ability to solve MaRs-IB puzzles was moderately predictive of their performance on a working memory task and the ICAR matrix reasoning test, indicating satisfactory convergent validity. At the conclusion of this study, \cite{chierchia2019matrix} made publicly available a summary of the difficulty (i.e. proportion correct responses) of each item template and clone in the MaRs-IB so that other researchers could construct their own test measures of custom difficulty and duration.

Two important limitations in the experiment conducted by \cite{chierchia2019matrix} undermine the utility of these summary statistics, and therefore impede the design of new matrix reasoning tasks using the MaRs-IB. The first is that, given the fixed-order task design,  there was relatively few responses collected for the majority of the items. Indeed only half (42 of 80 items) were completed by 100 or more participants. As such, one cannot be confident about the true difficulty of a majority of the items (an issue that is exacerbated when considering that each item has six clones). A second issue is participants were able to choose for themselves whether to prioritize accuracy (number of correct responses) or productivity (number of items reached). As a consequence, response data for items that came later in the fixed-order task are increasingly likely to have been produced by participants sacrificing accuracy for speed, thereby biasing the proportion correct measure for these items (see the supplementary materials for evidence of this effect). Therefore, additional investigation of the functioning and psychometric properties of the MaRs-IB is warranted.

Here we provide a more extensive study and psychometric validation of the MaRs-IB. The most important contribution of the current work is the characterization of (almost) every item in the MaRs-IB using item response theory (IRT; \citealt{embretson2013item, de2013theory}). IRT models confer several important advantages for the purposes of psychometric validation. First they provide an estimate of the \textit{difficulty} and \textit{discrimination} of each item, where the latter is an index of how much information an item provides about ability. Using these quantities, IRT models in turn allow experimenters to compute the test information function (TIF), which quantifies the reliability of a test in measuring performance across a range of ability levels. Finally, IRT models make possible optimal test assembly  \citep{van1998optimal}, or the design of new tests with maximal reliability given experimenter-defined constraints (e.g. test duration and/or difficulty).

A second contribution provided here is an analysis of how item properties shape the psychometric functioning of items in the MaRs-IB. Using explanatory item response modeling \citep{de2004explanatory, wilson2008explanatory}, we investigate how item complexity, i.e. the number of elements and number of rules, determine item functioning (i.e. difficulty, discrimination). We modeled these two item attributes because they have been previously identified as among the strongest predictors of item difficulty in matrix reasoning tasks \citep{embretson1998cognitive, primi2001complexity}. Thus this explanatory analysis provides another means by which to validate the success of the MaRs-IB design; that is, to confirm whether the MaRs-IB exhibits similar relationships between item properties and functioning as observed in other established matrix reasoning measures. As a secondary benefit, explanatory IRT models yield more accurate estimates of item parameters \citep{neuhaus2006separating}, which in turn helps to ensure that new MaRs-IB tests, designed using the item parameter estimates from these IRT models, function as expected.

A third contribution is an investigation of the exchangeability of item clones in the MaRs-IB, or if clones are (approximately) psychometrically equivalent. Establishing whether the MaRs-IB possesses this property is critical for the design of new matrix reasoning tasks using these items. If item clones are exchangeable, then new parallel test forms of equivalent difficulty and discrimination can be generated simply by swapping out one set of item clones for another (e.g. exchanging clones that differ only in shape set). To test for exchangeability in the MaRs-IB, we use additive multilevel item structure (AMIS) models \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. These models enable us to quantify the variability in item functioning across clones by distractor type and shape set, which must be negligible if clones are to be treated as exchangeable.

The remainder of the paper proceeds as follows. First we report a calibration study in which a large sample of participants each completed a subset of items from the MaRs-IB. Using their response data, we then fit a series of multilevel item structure models in order to estimate item parameters for each item (and their clones). Using these same models, we also interrogate the relationship between item attributes and functioning within and cross items in the MaRs-IB. Then we report a second validation study, in which a smaller sample of participants completed one of three new MaRs-IB short-form tests. These tests --- designed using optimal test assembly methods applied to the item parameters estimated during the calibration study --- were found to have good psychometric properties and convergent validity. This second study provides a blueprint for how to construct new MaRs-IB test measures using the item parameter estimates that have been released as part of this study.

All data, code, and model outputs (including the item parameter estimates) are publicly available at: \url{https://github.com/ndawlab/mars-irt}.

\clearpage
\section{Calibration Study}

@@

The purpose of the first study was to calibrate the items in the MaRs-IB using response data collected from a large number of adult participants from the general population. In particular, we sought to accomplish the following three aims: (1) to quantify the psychometric properties (i.e. difficulty, discrimination) of the items in the MaRs-IB using item response models; (2) to measure the associations between item complexity and functioning; and (3) to determiner whether item clones in the MaRs-IB are exchangeable (i.e. psychometrically equivalent).

@@

In order to ensure sufficient sampling of every item template and clone in the MaRs-IB, participants were administered a pseudorandomly-selected set of 16 out of 64 total items.\footnote{We did not collect new data for the 16 easiest items in the MaRs-IB (items 1, 2, 3, 4, 5, 7, 8, 9, 32, 33, 38, 41, 43, 48, 57, 68) because performance on these items was at or near ceiling.} Item sets were constructed as follows: We subdivided the 64 items into 16 sets of four based on their dimensionality (a measure of item complexity defined in \citealt{chierchia2019matrix}). Participants were randomly assigned one item from each of the 16 sets. As a consequence, all participants received test forms of roughly equal difficulty. Importantly we also counterbalanced the assignment of item clones across participants, such that we had an approximately uniform number of responses available for each item by shape set (1, 2, or 3) and distractor type (minimal difference, MD, or or paired difference, PD; see \citealt{chierchia2019matrix} for details).

@@

In our analyses we considered two item template (level 1) attributes, element number and rule number, which have previously been identified as key determinants of item difficulty in matrix reasoning tasks \citep{embretson1998cognitive, primi2001complexity}. Element number refers to the number of geometric shapes in each cell of a matrix, whereas rule number refers to the number of relationships that govern the changes of these shapes across cells. In the MaRs-IB there are four principal rules: size change, color change, position change, and shape change. Across item templates, element number ranges from one to four (median = 2) and rule number ranges from one to six (median = 3). Element and rule number are uncorrelated across item templates ($\rho$ = 0.176), and were determined for each item via manual annotation.

At the clone-level (level 2), we modeled an additional two attributes: distractor type and mean response time. Distractor type refers to whether a clone uses MD-type versus PD-type distractors, whereas mean response time refers to the average time participants spent deliberating on that item clone. We note that, in contrast to all other modeled attributes, average response time is not an intrinsic property of an item. We elected to include it in our models, however, because including response time as a covariate has previously been shown to improve parameter estimation \citep{bertling2018using}. To preclude any collinearity between the attributes, mean response time was orthogonalized with respect to the three other item attributes; as such, any effects of response time reflects residual structure after accounting for all other item attributes.

We deliberately chose not to incorporate shape set (i.e. 1, 2, or 3) as a clone-level attribute for the following reasons. As described above, every item template in the MaRs-IB has three unique versions that differ only in the geometric shapes populating its cells. There are 45 unique geometric shapes in total, organized into nine stimulus sets, where each item clone draws a subset of shapes from one of the nine sets (see Figure S2 for examples). As such, shape set --- indicating whether a clone is the 1st, 2nd, or 3rd version of an item template --- is not a meaningful label and we therefore do not incorporate it into the models. Furthermore, item clones drawing from the same shape set are not always populated by the same geometric shapes, making complicated the possibility of modeling the nine stimulus sets instead. The most rigorous way of modeling the presence or absence of a given geometric shape on item functioning would be to include each as a binary attribute predicting item difficulty and discrimination. In order to keep the models simple, we elected to leave shape set unmodeled and accounted for by the residual variability terms instead.

@@

The AMIS modeling framework provides a natural means of accomplishing the three objectives of the calibration study. Through all of the models above, we obtain estimates of the the psychometric properties (i.e. difficulty, discrimination) of every item template and clone in the MaRs-IB (Aim 1). Through the level-1 fixed effects, we are able to test for associations between item complexity (i.e. element and rule number) and item functioning (Aim 2). We predicted that both attributes would be positively associated with item difficulty; that is, all else equal, items with either a greater number of elements and rules would be more difficult. In contrast, we had no hypotheses about the impact of these attributes on item discrimination.

Finally, through both the level-2 fixed and random effects, we can determine whether item clones in the MaRs-IB are exchangeable (Aim 3). For example, an association between distractor type and either item parameter would entail that clones are not psychometrically equivalent. Moreover, if either level-2 residual variance term were non-negligible, that would similarly entail that not all item clones are equivalent due to one or more unmodeled features (e.g. low-level perceptual properties).

@@

In each of the six models described above, we incorporated person-level attributes as predictors of latent ability ($\theta_i$) following an explanatory item response modeling approach \citep{wilson2008explanatory}:

\begin{equation} \label{eq:4}
\theta_i = \sum_{p=1}^P X_{ip} \rho_p + \epsilon_i    
\end{equation}

@@

We estimated person-level abilities as a function of four covariates: age, gender, mean response time, and $\Delta$ response time. This last term reflects the degree of change in a participant's deliberation time as a function of item difficulty (defined as one minus the proportion of correct responses for that item). Mean response time and $\Delta$ response time were calculated for each participant simultaneously through linear regression, where a given participant's (log-transformed) response times were regressed against an intercept and item difficulty. All covariates were standard-scored prior to analysis with the exception of gender, which was binary coded (male = -0.5, female = 0.5).

@@

The $\chi^2_{NC}$ discrepancy statistic is a measure of item fit based on the comparison of observed and expected proportions of participants at each score level (here we ignore scores of zero and one, which comprised 7 total participants). A global measure of fit at the test level is obtained by summing the discrepancy values over the groups. In turn, SGDDM measures the mean absolute conditional correlation between all pairs of items; that is, the SGDDM is an index of residual inter-item correlations unexplained by the model. The SGDDM is important in this context as it tests for local dependence across items, the presence of which might otherwise lower the efficiency of future test forms via the inclusion of nonindependent or redundant items. For both discrepancy statistics, the PPP value is the proportion of draws in which the posterior predictive discrepancy is equal to or higher than the realized discrepancy. A poor model fit to the data is indicated when the PPP-values are extreme (PPP $\leq$ 0.05 or PPP $\geq$ 0.95).

@@

On average, participants completed 9.6 of 16 items correctly (sd = 3.1, IQR = 8.0 -- 12.0) (Figure \ref{fig:fig01}a). Timing out occurred on only 1.8\% of trials. A total of 43 participants (2.9\%) performed below chance (i.e. fewer than four items correct), while only 11 participants (0.7\%) solved all 16 of their items. Thus, over 95\% of observed scores were in a reasonable range. These results corroborate those reported by \cite{chierchia2019matrix} in that there were no obvious ceiling or floor effects in performance, and that the majority of participants were sufficiently motivated to participate in spite of the low-stakes testing environment.

@@

In sum, we found evidence of a speed-accuracy trade-off in performance in this sample. Accurate responses and more accurate participants were both slower overall. Notably, better-performing participants demonstrated adjustments to their response times as function of item difficulty; whereas the lowest scoring participants maintained the same work rate regardless of item difficulty, the highest scoring participants showed the largest slowing in responding as items became more challenging. We return to these results in the General Discussion.

@@

In the second wave of model fitting, in which models varied only in their specification of item discrimination, Model 5 demonstrated the best fit to the data. By comparison, Model 5 demonstrated smaller improvements in fit over Model 4 ($\Delta$ LOCO = 5.12, se = 4.07) and Model 6 ($\Delta$ LOCO = 3.29, se = 1.84). This result suggests the presence of non-negligible residual variance in item discrimination across item templates but not clones. It must be noted these differences in LOCO values from each model to the next were each within two standard errors of the mean, indicating only weak predictive improvement of Model 5 over the others \citep{vehtari2022cv}. As such we will select Model 5 as the best-fitting model overall, but proceed in discussing it with caution. We also note that all second-wave models exhibited better fits to the data than the first-wave models, supporting the inclusion of item discrimination parameters.

@@

Together element number and rule number explained approximately 67.6\% of the variance in difficulty across the 64 item templates. In other words, the majority of variability in difficulty across item templates can be attributed to their complexity. Individually the number of elements and number of rules explain 29.5\% and 40.6\% of the variance in item template difficulty, respectively. Across all 384 item clones, these two attributes explain 38.6\% of the variance in difficulty.

@@

Contrary to the assumption that item clones are exchangeable, we found an association between distractor type and item difficulty ($\delta_{\beta_3}$ = 1.105, 95\% HDI = 0.940 -- 1.269). Items with MD distractors were associated with an estimated 19.9\% reduction in accuracy compared to their equivalent items with PD distractors. Together element number, rule number, and distractor type explained slightly over half (52.0\%) of the variance in item difficulty. Despite this item response models that included a clone-level residual variance term for item difficulty was preferred to models without. Thus, the results indicate that not all item clones are not equally difficult and are thus cannot be assumed to be exchangeable.

To better demonstrate the residual variance in item difficulty across clones, the estimated item difficulty per clone organized by distractor type and shape set is presented in Figure \ref{fig:fig02}. The relative increase in difficulty for items with MD-type distractors compared to PD-type distractors is visually apparent (Figure \ref{fig:fig02}a). In contrast, the pattern of item difficulties across clones by shape set is more complicated (Figure \ref{fig:fig02}b). After controlling for distractor type and mean response time, the magnitude of the clone-level residual variance is $\epsilon_{\beta k}$ = 0.620 (95\% HDI = 0.538 -- 0.701). Across the 128 unique sets of clones (i.e. across item and distractor type), that translates into an average difference in difficulty of 0.752 (95\% HDI = 0.630 -- 0.877) between the easiest and hardest item clones of each set. In other words, the magnitude of residual variance in difficulty across clones is roughly two-thirds as large as the effect of distractor type. Importantly, the most difficult clone in each triad varied almost uniformly by shape set (shape set 1 = 39; shape set 2 = 32; shape set 3 = 57). Thus, the sources of residual variability in item difficulty across clones is complex and likely reflect the idiosyncratic contributions of low-level perceptual features.

>>>>

\abstract{Matrix reasoning tasks are among the most widely used measures of cognitive ability in the behavioral sciences, but the paucity of matrix reasoning tests in the public domain complicates their use. Here we present an extensive investigation and psychometric validation of the matrix reasoning item bank (MaRs-IB), an open-access set of matrix reasoning items. In a first study, we calibrate the psychometric functioning of the items in the MaRs-IB in a large sample adults participants (N=1501). Using additive multilevel item structure models, we establish that the MaRs-IB has many desirable psychometric properties: its items span a wide range of difficulty, possess medium- to-large levels of discrimination, and exhibit robust associations between item complexity and difficulty. However, we also find that item clones are not always psychometrically equivalent and cannot assumed to be exchangeable. In a second study, we demonstrate how researchers can use the item parameters we estimated to design new MaRs-IB tests using optimal item assembly. With these methods, we design three parallel short-form measures that exhibited good score reliability and convergent validity in a second sample of adults (N=300). We hope that the materials, results, and tutorials made available here will encourage researchers to use the MaRs-IB in their studies.}

@@

Matrix reasoning tasks are among the most widely used measures of cognitive ability in the behavioral sciences. Much of their popularity undoubtedly reflects their versatility of use. Matrix reasoning tasks are strong (albeit impure) indicators of general intelligence \citep{gignac2015raven} and working memory capacity \citep{kane2004generality, unsworth2005working}. They are predictive of important real-world outcomes such as childhood academic achievement \citep{roth2015intelligence} and performance on college entrance exams \citep{frey2004scholastic, koenig2008act}. In low-stakes testing settings (i.e. where participants incur little or no cost for poor performance), matrix reasoning tasks can additionally function as measures of motivation, willingness to expend mental effort, and other facets of personality \citep{duckworth2011role, gignac2019maximum}. In studies of psychiatric populations, performance on matrix reasoning tasks have also been used to control for general disruptions to cognitive ability when specific domains of cognition are of primary interest \citep{gillan2016characterizing, rouault2018psychiatric, moutoussis2021decision}.

Unfortunately there are a number of obstacles to using matrix reasoning tasks as part of behavioral research. One challenge is the problem of copyright. Many of the most prominent matrix reasoning tasks, such as the WAIS and WASI matrix reasoning subtests \citep{wechsler1999wechsler, corporation2008wechsler}, are not free to use and have legal restrictions against digitization. A second challenge is that, across all of the matrix reasoning tests in the public domain, there are relatively few unique items available. The Hagen matrices test \citep{heydasch2014hagen} and ICAR matrix reasoning test \citep{condon2014international}, for example, have only 20 and 16 items respectively. The availability of only a limited number of items raises the possibility of repeat exposure effects, which threaten the validity of these measures  \citep{ng1974applicability, bors2003effect}. This problem is exacerbated in the current era of online experiments, where multiple groups of researchers may be inadvertently administering the same test to the same participants recruited from the same online labor platforms.

Noting these challenges, \cite{knoll2016window} / \cite{chierchia2019matrix} developed and made publicly available the matrix reasoning item bank (MaRs-IB), a collection of 80 matrix reasoning puzzles. Each puzzle in the MaRs-IB consists of a 3x3 matrix containing geometric shapes in eight out of nine cells (Figure \ref{fig:fig00}). Across cells, some or all of these shapes change according to a number of abstract rules. Based on these rules, participants must deduce which of four response options correctly completes the matrix. The puzzles in the MaRs-IB vary in their complexity, both in the number of elements per cell and number of rules determining the relations of these elements across cells (Figure \ref{fig:fig00}a/b). To increase the reusability of the MaRs-IB, each puzzle (hereafter referred to as an item template) has six clones. A single item's clones possess identical structural properties (i.e. number of elements, number of rules) but may vary in their distractors (two unique sets per template) or perceptual features (three unique shape sets per template; Figure \ref{fig:fig00}c/d). Thus, the MaRs-IB addresses several of the issues associated with other matrix reasoning tests.

\cite{chierchia2019matrix} conducted an initial study of the psychometric properties of the MaRs-IB in which a sample of 659 adult, adolescent, and child participants had eight minutes to complete as many MaRs-IB items as possible. Of note, items were presented in the same order to all participants. The authors found that performance on this task had good split-half and test-retest reliability. Moreover, they found that participants' ability to solve MaRs-IB puzzles was moderately predictive of their performance on a working memory task and the ICAR matrix reasoning test, indicating satisfactory convergent validity. At the conclusion of this study, \cite{chierchia2019matrix} made publicly available a summary of the difficulty (i.e. proportion correct responses) of each item template and clone in the MaRs-IB so that other researchers could construct their own test measures of custom difficulty and duration.

Two important limitations in the experiment conducted by \cite{chierchia2019matrix} undermine the utility of these summary statistics, and therefore impede the design of new matrix reasoning tasks using the MaRs-IB. The first is that, given the fixed-order task design,  there was relatively few responses collected for the majority of the items. Indeed only 42 of 80 item templates were completed by 100 or more participants. As such, one cannot be confident about the true difficulty of a majority of the items (an issue that is exacerbated when considering that each template has six clones). A second issue is participants were able to choose for themselves whether to prioritize accuracy (number of correct responses) or productivity (number of items reached). As a consequence, response data for items that came later in the fixed-order task are increasingly likely to have been produced by participants sacrificing accuracy for speed, thereby biasing the proportion correct measure for these items (see the supplementary materials for evidence of this effect). Therefore, additional investigation of the functioning and psychometric properties of the MaRs-IB is warranted.

Here we provide a more extensive study and psychometric validation of the MaRs-IB. The most important contribution of the current work is the characterization of (almost) every item in the MaRs-IB using item response theory (IRT; \citealt{embretson2013item, de2013theory}). IRT models confer several important advantages for the purposes of psychometric validation. First they provide an estimate of the \textit{difficulty} and \textit{discrimination} of each item, where the latter is an index of how much information an item provides about ability. Using these quantities, IRT models in turn allow experimenters to compute the test information function (TIF), which quantifies the reliability of a test in measuring performance given a level of ability. Finally, IRT models make possible optimal test assembly  \citep{van1998optimal}, or the design of new tests with maximal reliability given researcher-specified constraints (e.g. test duration or difficulty).

A second contribution provided here is an analysis of how item properties shape the psychometric functioning (i.e. difficulty, discrimination) of items in the MaRs-IB. Using explanatory item response modeling \citep{de2004explanatory, wilson2008explanatory}, we specifically investigate how item complexity, i.e. the number of elements and number of rules, determine item functioning. We modeled these two item attributes because they have been previously identified as among the strongest predictors of item difficulty in matrix reasoning tasks \citep{embretson1998cognitive, primi2001complexity}. This explanatory analysis provides another means by which to validate the design of the MaRs-IB; that is, to determine whether the MaRs-IB exhibits the same associations between item properties and functioning as found in other established matrix reasoning tests. As a secondary benefit, explanatory IRT models yield more precise estimates of item parameters \citep{neuhaus2006separating}, which in turn help to ensure that any new MaRs-IB tests designed based on those estimates function as expected.

A third contribution is an investigation of the exchangeability of item clones in the MaRs-IB, or if clones are (approximately) psychometrically equivalent. Establishing whether the MaRs-IB possesses this property is critical for the design of new matrix reasoning tasks using these items. If item clones are exchangeable, then new parallel test forms of equivalent difficulty and discrimination can be generated simply by substituting a an item clone from one shape set for another. To test for exchangeability, we use additive multilevel item structure (AMIS) models \citep{geerlings2011modeling, cho2014additive, lathrop2017item}. These models enable us to quantify the variability in item difficulty and discrimination across clones, by distractor type and shape set, which must be negligible if clones are to be treated as exchangeable.

The remainder of the paper proceeds as follows. First we report a calibration study in which a large sample of adult participants completed a subset of items from the MaRs-IB. Using their response data, we then fit a series of multilevel item structure models in order to estimate item parameters for each item (and their clones). Using these same models, we also interrogate the relationship between item attributes and functioning items in the MaRs-IB. Then we report a second validation study, in which a smaller sample of participants completed one of three new MaRs-IB short-form tests. These tests --- designed using optimal test assembly methods applied to the item parameters estimated during the calibration study --- were found to have good psychometric properties and convergent validity. This second study provides a blueprint for how to construct new MaRs-IB test measures using the item parameter estimates that were derived in the calibration study.

All data, code, model outputs (including the estimated item parameters), and tutorials are publicly available at: \url{https://github.com/ndawlab/mars-irt}.

\section{Calibration Study}

@@

The purpose of the first study was to calibrate the items in the MaRs-IB using response data collected from a large number of adult participants recruited from the general population. In particular, we sought to accomplish the following three aims: (1) to quantify the psychometric properties (i.e. difficulty, discrimination) of the items in the MaRs-IB using item response models; (2) to measure the associations between item complexity and functioning; and (3) to determine whether item clones in the MaRs-IB are psychometrically equivalent and exchangeable.

@@

In order to ensure sufficient sampling of every item template and clone in the MaRs-IB, participants were administered a pseudorandomly-selected set of 16 out of 64 total items.\footnote{We did not collect new data for the 16 easiest items in the MaRs-IB (items 1, 2, 3, 4, 5, 7, 8, 9, 32, 33, 38, 41, 43, 48, 57, 68) because performance on these items was at or near ceiling.} Item sets were constructed as follows: We subdivided the 64 items into 16 sets of four based on their dimensionality (a measure of item complexity defined in \citealt{chierchia2019matrix}). Participants were randomly assigned one item from each of the 16 sets. As such, all participants received test forms of roughly equal difficulty. Importantly we also counterbalanced the assignment of item clones across participants, such that we had an approximately uniform number of responses available for each item by shape set (1, 2, or 3) and distractor type (minimal difference, MD, or paired difference, PD; see \citealt{chierchia2019matrix} for details).

@@

In our analyses we considered two item template (level 1) attributes: element number and rule number. These have previously been identified as key determinants of item difficulty in matrix reasoning tasks \citep{embretson1998cognitive, primi2001complexity}. Element number refers to the number of geometric shapes in each cell of a matrix, whereas rule number refers to the number of relationships that govern the changes among these shapes across cells. In the MaRs-IB there are four principal rules: size change, color change, position change, and shape change. Across item templates, element number ranges from one to four (median = 2) and rule number ranges from one to six (median = 3). Element and rule number are uncorrelated across item templates ($\rho$ = 0.176), and were determined for each item via manual annotation.

At the clone-level (level 2), we modeled an additional two attributes: distractor type and mean response time. Distractor type refers to whether a clone uses MD-type versus PD-type distractors, whereas mean response time refers to the average time participants spent deliberating on that item clone. We note that, in contrast to the other modeled attributes, average response time is not an intrinsic property of an item. We elected to include it in our models, however, because including response time as a covariate has previously been shown to improve parameter estimation \citep{bertling2018using}.\footnote{Removing mean response time as a clone-level attribute did not appreciably change the estimated item parameters; see Figure S1.} To preclude any collinearity between the attributes, mean response time was orthogonalized with respect to the three other item attributes; as such, any effects of response time reflects residual structure after accounting for all other item attributes.

We deliberately chose not to incorporate shape set (i.e. 1, 2, or 3) as a clone-level attribute. The allocation of geometric shapes to item clones in the MaRs-IB is complex, such that not all clones of the same shape set share the same perceptual elements (see the supplementary materials for a full discussion). In other words, shape set does not constitute a meaningful nominal variable and was therefore not included in the model. Instead, the contribution of perceptual elements to item functioning is captured by residual variability terms of the item structure models.

@@

The AMIS modeling framework provides a natural means of accomplishing the three objectives of the calibration study. Via all of the models above, we obtain estimates of the the psychometric properties (i.e. difficulty, discrimination) of every item template and clone in the MaRs-IB (Aim 1). Through the level-1 fixed effects, we are able to test for associations between item complexity (i.e. element and rule number) and item functioning (Aim 2). We predicted that both attributes would be positively associated with item difficulty; that is, all else equal, items with either a greater number of elements and rules would be more difficult. In contrast, we had no hypotheses about the impact of these attributes on item discrimination.

Finally, through both the level-2 fixed and random effects, we can determine whether item clones in the MaRs-IB are exchangeable (Aim 3). For example, a credible association between distractor type and either item parameter would entail that clones are not psychometrically equivalent. Moreover, if either level-2 residual variance term were non-negligible, that would similarly entail that not all item clones are equivalent due to unmodeled attributes (e.g. low-level perceptual features).

@@

In each of the six models described above, we incorporated person-level attributes as predictors of latent ability ($\theta_i$) following an explanatory item response modeling approach \citep{wilson2008explanatory}: 
\begin{equation} \label{eq:4}
\theta_i = \sum_{p=1}^P X_{ip} \rho_p + \epsilon_i    
\end{equation}

@@

We estimated person-level abilities as a function of four covariates: age, gender, mean response time, and $\Delta$ response time. This last term reflects the degree of change in a participant's response time as a function of item difficulty (defined as one minus the proportion of correct responses for that item), and was included because of its relationship to participants' performance (see Results). Mean response time and $\Delta$ response time were calculated for each participant simultaneously via linear regression, where a given participant's (log-transformed) response times were regressed against an intercept and item difficulty. All covariates were standard-scored prior to analysis with the exception of gender, which was binary coded (male = -0.5, female = 0.5).

@@

The $\chi^2_{NC}$ discrepancy statistic is a measure of item fit based on the comparison of observed and expected proportions of participants at each score level (here we ignore scores of zero and one, which comprised 7 total participants). A global measure of fit at the test level is obtained by summing the discrepancy values over the groups. In turn, SGDDM measures the mean absolute conditional correlation between all pairs of items; that is, the SGDDM is an index of residual inter-item correlations unexplained by the model. The SGDDM is important in this context because it tests for local dependence across items, the presence of which might otherwise lower the efficiency of future test forms by including nonindependent or redundant items. For both discrepancy statistics, the PPP value is the proportion of draws in which the posterior predictive discrepancy is equal to or higher than the realized discrepancy. A poor model fit to the data is indicated when the PPP-values are extreme (PPP $\leq$ 0.05 or PPP $\geq$ 0.95).

@@

On average, participants completed 9.6 of 16 items correctly (sd = 3.1, IQR = 8.0 -- 12.0) (Figure \ref{fig:fig01}a). Timing out occurred on only 1.8\% of trials (these were subsequently coded as incorrect responses). A total of 43 participants (2.9\%) performed below chance (i.e. fewer than four items correct), while only 11 participants (0.7\%) solved all 16 of their items. Thus, over 95\% of observed scores were in a reasonable range. These results corroborate those reported by \cite{chierchia2019matrix} in that there were no obvious ceiling or floor effects in performance, and that the majority of participants were sufficiently motivated to participate in spite of the low-stakes testing environment.

@@

In sum, we found evidence of a speed-accuracy trade-off in performance in this sample. Accurate responses and more accurate participants were both slower overall. Notably, better-performing participants demonstrated adjustments to their response times as function of item difficulty; whereas the lowest scoring participants maintained an equivalent work rate regardless of item difficulty, the highest scoring participants showed the largest slowing in responding as items became more challenging. We return to these results in the General Discussion.

@@

In the second wave of model fitting, in which models varied only in their specification of item discrimination, Model 5 demonstrated the best fit to the data. In contrast, Model 5 yielded smaller improvements in fit over Model 4 ($\Delta$ LOCO = 5.12, se = 4.07) and Model 6 ($\Delta$ LOCO = 3.29, se = 1.84). This result suggests the presence of non-negligible residual variance in item discrimination across item templates but not clones. It must be noted these differences in LOCO values from each model to the next were each within two standard errors of the mean, indicating only weak predictive improvement of Model 5 over the others \citep{vehtari2022cv}. As such we will select Model 5 as the best-fitting model overall, but proceed in discussing it with caution. We also note that all second-wave models exhibited better fits to the data than the first-wave models, supporting the inclusion of item discrimination parameters.

@@

Together element number and rule number explained approximately 67.6\% of the variance in difficulty across the 64 item templates. That is, the majority of variability in difficulty across item templates can be attributed to their complexity. Individually the number of elements and number of rules respectively explain 29.5\% and 40.6\% of the variance in item template difficulty. Across all 384 item clones, these two attributes explain 38.6\% of the variance in difficulty.

@@

Contrary to the assumption that item clones are exchangeable, we found an association between distractor type and item difficulty ($\delta_{\beta_3}$ = 1.105, 95\% HDI = 0.940 -- 1.269). Items with MD distractors were associated with an estimated 19.9\% reduction in accuracy compared to their equivalent items with PD distractors. Together element number, rule number, and distractor type explained slightly over half (52.0\%) of the variance in item difficulty. Despite this, an item response model that included a clone- level residual variance term for item difficulty was preferred to one without. In sum, the results clearly indicate that item clones are not equally difficult and cannot be assumed to be exchangeable.

To make clear the residual variance in item difficulty across clones, the estimated item difficulty per clone is presented in Figure \ref{fig:fig02}. The relative increase in difficulty for items with MD-type distractors compared to PD-type distractors is visually apparent (Figure \ref{fig:fig02}a). In contrast the pattern of residual difficulty across clones, after accounting for all modeled sources of variance, is more complicated (Figure \ref{fig:fig02}b). The magnitude of the clone-level residual variance is $\epsilon_{\beta k}$ = 0.620 (95\% HDI = 0.538 -- 0.701), or 35.1\% of the total variance in difficulty across item clones. This corresponds to a mean absolute difference in difficulty of 0.705 (95\% HDI = 0.612 -- 0.786) across clones per item and distractor type, or roughly two-thirds as large as the effect of distractor type. Thus, the residual variability in item difficulty across clones is large, complex, and likely reflects the idiosyncratic contributions of low-level perceptual features.
